---
title: "Gov 50 Final Project"
author: "Rebecca Buehler"
description: "A Gender-Based Analysis of Violence in the Tigray War"
output:
  distill::distill_article:
    self_contained: false
---

## Final Report

## Introduction

One year ago, the war in Tigray was the largest violent conflict in the world. However, based on its coverage in the media, one never would have known. By the end of 2022, two years after the conflict began, researchers from Ghent University estimated that as many as 600,000 civilians had died: between 50,000 and 100,000 had been killed; 150,000 were victims of famine; and more than 100,000 others had died due to the lack of medical aid (Green). For reference, as of November 2023, the UN estimates the war in Ukraine to have killed 10,000 civilians (source). Yet, while the war in Ukraine has received daily attention from newspapers, governments, and organizations around the world, the war in Tigray appears forgotten by the West (Green).

## Research Question #1

In her 2012 disaggregated analysis of violence against civilians in African civil wars, Clionadh Raleigh, the Executive Director of the Armed Conflict Location and Event Data (ACLED) project, recognizes the tension between the popular accounts that civilians represent 80-90% of all conflict deaths and conclusions that far fewer civilians are killed during civil wars (Raleigh 2012, Goldstein 2011). 

The staggering estimates of civilian casualties in Tigray would suggest support for the first hypothesis. However, these estimates do not compare civilian casualties to those for combatant fatalities (source). In the following data analysis, I will evaluate the casualty data from the Tigray War against the popular hypothesis that 80-90% of all conflict deaths are civilians and Goldstein's counterargument that this proportion is much lower in civil conflicts. I will use data from ACLED, the most comprehensive source of subnational political violence data, to answer the following research question: What proportion of lethal violence during the Tigray War targeted civilians? I hypothesize that the proportion of civilian fatalities will be significantly higher than the proportion of combatant fatalities. 

## Research Question #2 

A secondary tension relevant to the Tigray War occurs between the popular tendency to view "women and children" as the primary victims of war and the reality of gendered civilian victimization (source). This assumption of female vulnerability is, in part, the product of deeply ingrained gender stereotypes, media coverage that centers stories and images of women and children, and the humanitarian practice of identifying, prioritizing, and evacuating "women and children" as a particularly vulnerable group (Carpenter 2006). It is true that women confront unique vulnerabilities in conflict contexts and their aftermath. However, in conflict contexts around the world, civilian men are twice as likely to die compared to civilian women (Ormhaug et al. 2009). Men made up 82% of civilians killed in the Iraq War from 2003-2005, nearly 90% of civilian casualties in the Syrian civil war from 2011-2015, and more than 80% of civilian deaths and disappearances in the Timor-Leste and Sierra Leone conflicts (Ormhaug et al. 2009, Syrian Observatory 2023, Silva and Ball 2006, Conibere 2004). Available evidence suggests that women suffer disproportionate rates of conflict-related sexual violence and more long-term consequences of conflict (Cohen and Nordås 2014, Plümper and Neumayer 2006). However, men are more likely to suffer from nearly every form of conflict-related violence, including every form of lethal violence (Carpenter 2006).

Although ACLED does not provide gender-disaggregated casualty data for Ethiopia, its data set notes when violence that specifically targeted women (without doing the same for violence that targeted men, an example in and of itself of the general tendency to overlook male victimization). Using this data, I will seek to answer the following research question: What proportion of lethal violence against civilians during the Tigray War targeted women? I hypothesize that the proportion of lethal violence that specifically targets women will not constitute the majority of lethal violence against civilians, contrary to the popular conception of women as the primary victims of war. In the following data visualization, I will present these findings. 

## Data analysis

The ACLED dataset is limited due to the difficulty of obtaining casualty data from the Tigray region, and it does not represent the total number of civilian combatant deaths during the war (Tigray Conflict Coding Methodology). 

Region and country were already held constant. I controlled for region (Afar, Amhara, Tigray), year of the conflict (2020-2023)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(ggplot2)
library(dplyr)
library(stargazer)
library(anytime)
library(lubridate)
library(arm)
```

```{r}
ethiopia <- read_excel("~/Documents/Thesis/Data/ACLED/Ethiopia_1997-2023_Sep22.xlsx") 

tigray_war <- ethiopia |>
  filter(
    EVENT_DATE >= "2020-11-03",
    ADMIN1 == "Tigray" | (ADMIN1 == "Afar" & ACTOR1 == "TPLF: Tigray People's Liberation Front") | (ADMIN1 == "Amhara" & ACTOR1 == "TPLF: Tigray People's Liberation Front"))

tigray_war$EVENT_TYPE <- as.factor(tigray_war$EVENT_TYPE)

model1 <- lm(FATALITIES ~ EVENT_TYPE + YEAR + ADMIN1, data = tigray_war)

stargazer(model1, 
          title = "Event Type's Effect on Fatality Count", 
          type = "text",
          covariate.labels = c("Explosions/Remote Violence", "Protests",
                               "Riots","Strategic Developments",
                               "Violence Against Civilians",
                               "Year"),
          dep.var.labels = "Fatalities")
```

This analysis looks at the effect of event type (Battles, Explosions/Remote Violence, Protests, Riots, Strategic Developments, Civilian Violence) and year of the conflict (2020-2023) on the number of fatalities. Sub-events categorized as Civilian Violence include attacks, sexual violence, and abductions/forced disappearances. This analysis includes data from 4,901 events between the start of the Tigray War on November 3, 2020 and September 21, 2023. This data was collected by the Armed Conflict Location & Event Data (ACLED) project and published by the Ethiopia Peace Observatory.

A regression of event type on fatalities with “Battles” as the baseline event type shows that violence against civilians has an associated increase in fatalities of 9.09. One would expect roughly 2 more people to die in a given event targeting civilians than in a battle. The standard error is 0.5926, and these results are statistically significant.
## Add interpretation of year and admin1

## Data visualization

```{r}
civilian_violence <- ethiopia |>
  filter(
    CIVILIAN_TARGETING == "Civilian targeting", 
    EVENT_DATE >= "2020-11-03",
    ADMIN1 == "Tigray" | (ADMIN1 == "Afar" & ACTOR1 == "TPLF: Tigray People's Liberation Front") | (ADMIN1 == "Amhara" & ACTOR1 == "TPLF: Tigray People's Liberation Front")) |>
  mutate(
    WOMEN = ifelse(grepl("Women", ASSOC_ACTOR_2) == TRUE, 1, 0),
    month = format(as.Date(EVENT_DATE, "%Y-%m-%d"), "%Y-%m"),
    year = format(as.Date(EVENT_DATE, "%Y-%m-%d"), "%Y"))

civilian_violence_plot <- civilian_violence |>
  group_by(month) |>
  mutate(fatalities_per_month = sum(FATALITIES)) |> 
  filter(WOMEN == 1) |>
  mutate(women_fatalities_per_month = sum(FATALITIES)) |>
  ggplot() +
  facet_grid(~year, scales="free", space = "free_x", switch = "x") +
  geom_col(mapping = aes(x = month, y = fatalities_per_month, fill = 'Fatalities from violence targeting civilians')) +
  geom_col(mapping = aes(x = month, y = women_fatalities_per_month, fill = 'Fatalities from violence targeting women')) +
  theme(axis.text.x = element_blank(), 
        panel.spacing = unit(0.05, "lines"),
        strip.placement = "outside") +
  labs(
    x = "Year",
    y = "Fatalities per Month",
    title = "Civilian Casualities in the Tigray War from November 2020 to Present"
    ) +
  scale_fill_manual(name = 'Fatalities',
                     values = c('Fatalities from violence targeting civilians'='darkslateblue', 'Fatalities from violence targeting women'= 'pink'))
civilian_violence_plot
```


## Conclusion

The war in Tigray saw war crimes and crimes against humanity, including deliberate starvation, torture, ethnic cleansing, extra-judicial killings, and sexual violence (Green). Despite the cessation of hostilities in November 2022, the violence and suffering in the Tigray region continue today. With respect to these staggering statistics, the aid and attention from the rest of the world have been dramatically insufficient. There is a dire need for humanitarian assistance and post-conflict reconstruction. The international community has the obligation to respond.



